TacticToe: Learning to Prove with Tactics

### 2018-03-14 tactictoe_jv.tex
some fixes, in particular Def 10 seemed in need of flipping

\begin{definition} (Coverage)\\
Let $\mathfrak{T}$ be the database of tactic-goal pairs. We can define the
coverage $\textit{\mbox{Coverage}}(t)$ of a tactic $t$ by the number of times
this tactic
appears in
$\mathfrak{T}$. Expressing this in a formula, we get:
  \[\textit{\mbox{Coverage}}(t) =_{def} card\ \lbrace g\ |\ (t,g) \in
  \mathfrak{T} \rbrace \]
Intuitively, this notion estimates how general a tactic is by counting the
number of different goals it is useful for.
\end{definition}

\begin{definition} (Goal subsumption)\\
A goal subsumption $\le$ is a partial relation goals.
Assuming we have a way to estimate the number steps required to solve a goal,
a possibly useful subsumption definition could be:
\[g_1 \le g_2  \Leftrightarrow_{def} g_1 \mbox{ has a proof with a length
shorter
or equal to the proof of }
g_2\]
By default, we choose a minimal subsumption defined by:
\[g_1 \le g_2  \Leftrightarrow_{def} g_1 \mbox{ is }\alpha\mbox{-equivalent
to } g_2\]
\end{definition}

We can naturally extend any goal subsumption to a subsumption on
lists of goals.
\begin{definition} (Goal list subsumption)\\
A goal list subsumption is an partial relation on lists of goals.
Given two lists of goals $l_1$ and $l_2$, we can define it from goal
subsumption $\le$ by:
\[l_1 \le l_2  \Leftrightarrow_{def} \forall g_2 \in l_2.\ \exists g_1 \in l_1.\
g_1 \le g_2\]
\end{definition}

@@

If one of the tactic is failing then $t_1$ and $t_2$ are not comparable through
this relation.
Given the default subsumption on goals, $t_1$ subsumes $t_2$ on $g$ if and only
if $t_1(g)$ is a subset (modulo $\alpha$-equivalence) of $t_2(g)$.
\end{definition}

@@

All in all, a database build with the orthogonalization process contains
$(\mathbb{O}(t,g),g)$ instead of $(t,g)$.

\section{Abstraction}\label{sec:synthesis}
One of the major weakness of the previous version of \tactictoe was that
it could not create its own tactics. Indeed, sometimes no previous tactic is
adapted for a certain goal and creating a new tactic is necessary.
%Experiments in section, show how different degrees of tactic synthesis could
%improve the success rate of \tactictoe.
In this section we present a way to create tactics with different arguments
by abstracting them and re-instantiating them using a predictor for that kind
of argument. In the spirit of the orthogonalization method, we will try to
create tactics that are more general but have the same effect as the
ungeneralized one. The generalized tactics are typically slower that their
original variants, but the added flexibility is worthwhile in practice. Moreover,
since we impose a timeout on each tactic (0.05 seconds by default), very slow
tactics fail and thus are not selected.

@@

\paragraph{Reconstruction}
Tactic proofs produced by \tactictoe during this experiment are all verifiable
in \holfour. By the design of the proof search, reconstruction of \tactictoe
proof succeeds, unless one of the tactic modifies the state of \holfour in a
way that changes the behavior of a tactic used in the final proof. This has not
occurred a single time in our experiments.
More details on how a proof is extracted from the search tree is given in
Section~\ref{sec:proofdisplay}.
Comparatively, we achieve a reconstruction rate of 95 percent for \eprover,
by calling \metis for 2.0 seconds with the set of theorems used
in \eprover's proof as argument.

@@

\paragraph{Reinforcement learning}
All our feature vectors have been learned form human proofs. We now can now
also add tactic-goal pairs that appears in final \tactictoe's proof. To prevent
duplication of effort, orthogonalization of those
tactics is essential to have a beneficial effect.
Since recording and re-proving are intertwined during evaluation, the
additional data is available for the next proof search.
The hope is that the algorithm will improve faster by learning from its own
discovered proofs than from the human proof
scripts~\cite{DBLP:conf/cade/Urban07}. Side experiments show that
this one shot reinforcement learning method increases \tactictoe's success rate
by less than a percent.

@@

In order to appreciate the difficulty of the evaluation set from a human
perspective, the length distribution of human proofs in the
validation set is shown in Figure~\ref{fig:distrib}.  It is clear from the
graph that most of the human proofs are short. The proofs found by \tactictoe
follows a similar
distribution. This means that if \tactictoe finds a proof, there is about a 50
percent chance that it will be shorter than what a human would come up with.

@@

In Figure~\ref{fig:percentage}, we regroup the human proofs by their length to
see how well \tactictoe and \eprover performs on increased level of
difficulties. As expected, the longer the original proof is, the harder it is
for \tactictoe and \eprover to re-prove the theorem on their own.
The performance of \tactictoe is compelling with more than half of the theorems
that required a proof of length six being reproven. It is also consistently
better than \eprover for any proof length.

\section{Proof Recording}\label{sec:recording}
Recording proofs in an LCF-style proof assistant can be done at different
levels.
In \holfour all existing approaches relied on modifying the kernel. This was
used
either to export the primitive inference
steps~\cite{Wong95recordingand,DBLP:conf/itp/KumarH12}
or to record dependencies of theorems~\cite{tgck-cpp15}. This was not suitable
for our
purpose of learning proving strategies at the intermediate tactic level. We
therefore
discuss recording proofs in an LCF-style proof assistant, with the focus on
\holfour
in this section.
%
Rather than relying on the underlying programming language, we parse the
original theory file containing proof scripts. This enables us to extract the
code of each tactic.
Working with the string representation of a tactic is better than working with
its value:
\begin{itemize}
\item Equality between two functions is easy to compute from
their string representation, which allows us to avoid
predicting the same tactic repeatedly.
\item
The
results of
\tactictoe can be returned in a readable form to the user. In this way, the
user can learn from the feedback, analyze the proof and possibly improve on it.
Furthermore \tactictoe does not need to be installed to run
proof scripts generated by \tactictoe. In this way, the further development of
\tactictoe does not affect the robustness of \holfour.
\item
It is difficult (probably impossible) to transfer \sml values
between theories, since they are not run in the same \holfour session. In contrast,
\sml code can be exported and imported.
\item To produce a tactic value form a code is easy in \sml, which can be
achieved by using a reference to a tactic and update it with the function
\texttt{use}.
\end{itemize}

In order to transfer our tactic knowledge between theories, we want to be able
to re-use the code of a tactic recorded in one theory in an
other. We also would like to make sure that the code is interpretable and that
its interpretation does not change in the other theory.

>>>>

\begin{definition} (Coverage)\\
Let $\mathfrak{T}$ be the database of tactic-goal pairs. We define the
coverage $\textit{\mbox{Coverage}}(t)$ of a tactic $t$ by the number of times
this tactic
appears in
$\mathfrak{T}$. Expressing this in a formula, we get:
  \[\textit{\mbox{Coverage}}(t) =_{def} card\ \lbrace g\ |\ (t,g) \in
  \mathfrak{T} \rbrace \]
Intuitively, this notion estimates how general a tactic is by counting the
number of different goals it is useful for.
\end{definition}

\begin{definition} (Goal subsumption)\\
A goal subsumption $\le$ is a partial relation on goals.
Assuming we have a way to estimate the number of steps required to solve a goal,
a % possibly 
general and useful subsumption definition is for example:
% could be:
\[g_1 \le g_2  \Leftrightarrow_{def} g_1 \mbox{ has a proof with a length
shorter
or equal to the proof of }
g_2\]
By default, we however choose for efficiency reasons a minimal subsumption defined by:
\[g_1 \le g_2  \Leftrightarrow_{def} g_1 \mbox{ is }\alpha\mbox{-equivalent
to } g_2\]
\end{definition}

We can naturally extend any goal subsumption to a subsumption on
lists of goals.
\begin{definition} (Goal list subsumption)\\
A goal list subsumption is a partial relation on lists of goals.
Given two lists of goals $l_1$ and $l_2$, we define it from the goal
subsumption $\le$ by:
%\[l_1 \le l_2  \Leftrightarrow_{def} \forall g_2 \in l_2.\ \exists g_1 \in l_1.\
%g_1 \le g_2\]
\[l_1 \le l_2  \Leftrightarrow_{def} \forall g_1 \in l_1.\ \exists g_2 \in l_2.\
g_1 \le g_2\]
\end{definition}

@@

If one of the tactics is failing then $t_1$ and $t_2$ are not comparable through
this relation.
Given the default subsumption on goals, $t_1$ subsumes $t_2$ on $g$ if and only
if $t_1(g)$ is a subset (modulo $\alpha$-equivalence) of $t_2(g)$.
\end{definition}

@@

% All in all, 
A database built with the orthogonalization process hus contains
$(\mathbb{O}(t,g),g)$ instead of $(t,g)$.

\section{Abstraction}\label{sec:synthesis}
One of the major weaknesses of the previous version of \tactictoe was that
it could not create its own tactics. Indeed, sometimes no previous tactic is
adapted for a certain goal and creating a new tactic is necessary.
%Experiments in section, show how different degrees of tactic synthesis could
%improve the success rate of \tactictoe.
In this section we present a way to create tactics with different arguments
by abstracting them and re-instantiating them using a predictor for that kind
of argument. In the spirit of the orthogonalization method, we will try to
create tactics that are more general but have the same effect as the
% ungeneralized 
original one. The generalized tactics are typically slower that their
original variants, but the added flexibility is worthwhile in practice. Moreover,
since we impose a timeout on each tactic (0.05 seconds by default), very slow
tactics fail and thus are not selected.

@@

\paragraph{Reconstruction}
Tactic proofs produced by \tactictoe during this experiment are all verifiable
in \holfour. By the design of the proof search, reconstruction of \tactictoe
proof succeeds, unless one of the tactics modifies the state of \holfour in a
way that changes the behavior of a tactic used in the final proof. This has not
occurred a single time in our experiments.
More details on how a proof is extracted from the search tree is given in
Section~\ref{sec:proofdisplay}.
Comparatively, we achieve a reconstruction rate of 95 percent for \eprover,
by calling \metis for 2.0 seconds with the set of theorems used
in \eprover's proof as argument.

@@

\paragraph{Reinforcement learning}
All our feature vectors have been learned form human proofs. We now can now
also add tactic-goal pairs that appears in \tactictoe's final proof. To prevent
duplication of effort, orthogonalization of those
tactics is essential to have a beneficial effect.
Since recording and re-proving are intertwined during evaluation, the
additional data is available for the next proof search.
The hope is that the algorithm will improve faster by learning from its own
discovered proofs than from the human proof
scripts~\cite{DBLP:conf/cade/Urban07}. Side experiments show that
this one shot reinforcement learning method increases \tactictoe's success rate
by less than a percent.

@@

In order to appreciate the difficulty of the evaluation set from a human
perspective, the length distribution of human proofs in the
validation set is shown in Figure~\ref{fig:distrib}.  It is clear from the
graph that most of the human proofs are short. The proofs found by \tactictoe
follow a similar
distribution. % This means that 
If \tactictoe finds a proof, there is about a 50
percent chance that it will be shorter than what a human would come up with.

@@

In Figure~\ref{fig:percentage}, we regroup the human proofs by their length to
see how well \tactictoe and \eprover cope with increased levels of
difficulty. As expected, the longer the original proof is, the harder it is
for \tactictoe and \eprover to re-prove the theorem on their own.
The performance of \tactictoe is compelling with more than half of the theorems
that required a proof of length six being reproven. It is also consistently
better than \eprover for any proof length.

\section{Proof Recording}\label{sec:recording}
Recording proofs in an LCF-style proof assistant can be done at different
levels.
In \holfour all existing approaches relied on modifying the kernel. This was
used
either to export the primitive inference
steps~\cite{Wong95recordingand,DBLP:conf/itp/KumarH12}
or to record dependencies of theorems~\cite{tgck-cpp15}. This was not suitable
for our
purpose of learning proving strategies at the intermediate tactic level. We
therefore
discuss recording proofs in an LCF-style proof assistant, with the focus on
\holfour
in this section.
%
Rather than relying on the underlying programming language, we parse the
original theory file containing proof scripts. This enables us to extract the
code of each tactic.
Working with the string representation of a tactic is better than working with
its value:
\begin{itemize}
\item Equality between two functions is easy to compute from
their string representation, which allows us to avoid
predicting the same tactic repeatedly.
\item
The
results of
\tactictoe can be returned in a readable form to the user. In this way, the
user can learn from the feedback, analyze the proof and possibly improve on it.
Furthermore \tactictoe does not need to be installed to run
proof scripts generated by \tactictoe. In this way, the further development of
\tactictoe does not affect the robustness of \holfour.
\item
It is difficult (probably impossible) to transfer \sml values
between theories, since they are not run in the same \holfour session. In contrast,
\sml code can be exported and imported.
\item To produce a tactic value from a code is easy in \sml, which can be
achieved by using a reference to a tactic and updating it with the function
\texttt{use}.
\end{itemize}

In order to transfer our tactic knowledge between theories, we want to be able
to re-use the code of a tactic recorded in one theory in another. 
We also would like to make sure that the code is interpretable and that
its interpretation does not change in the other theory.
